Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-10-02 manuscript/main.tex
updated results

\documentclass[12pt, a4paper]{article}
\setcounter{tocdepth}{2}
\usepackage{graphicx} % Required for inserting images
\usepackage[parfill]{parskip} % Removes indent from new paragraphs
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[colorlinks = true,
			citecolor  = blue,
            linkcolor = black]{hyperref}
\usepackage[font={small,it}]{caption} % Figure caption font
\usepackage[labelfont=bf]{caption} % Bold figure x.x

@@

\section{Results}

    \subsection{Hamiltonian Monte Carlo}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/hmc_1k.png}
        \caption{Distribution of rank statistics for centered and non centered Hamiltoninan Monte Carlo (static parameters). Non centered histograms look approximately uniform. The centered parameterisation struggles to estimate $\sigma^2$}
    \end{figure}

    \subsection{Gaussian Mixture Approximation}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_1k.png}
        \caption{Distribution of rank statistics for centered and non centered Gaussian mixture model(static parameters). All parameters display a lack of calibration}
    \end{figure}

    \subsection{Reweigthing rank statistics of Gaussian Mixture}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/sir_1k.png}
        \caption{Distribution of rank statistics for weighted resampled model (static parameters) with 1000 SBC iterations. All parameters display a lack of calibration}
    \end{figure}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/sir_5k.png}
        \caption{Distribution of rank statistics for weighted resampled model (static parameters) with 5000 SBC iterations. Increasing number of SBC parameters does not improve results}
    \end{figure}

    \subsection{All models and parameters}
    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/dist_chisq.png}
        \caption{Distribution of chi squared statistics for all models.}
    \end{figure}

@@

>>>>

\documentclass[12pt, a4paper]{article}
\setcounter{tocdepth}{2}
\usepackage{graphicx} % Required for inserting images
\usepackage[parfill]{parskip} % Removes indent from new paragraphs
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[colorlinks = true,
			citecolor  = blue,
            linkcolor = black]{hyperref}
\usepackage[font={scriptsize}]{caption} % Figure caption font
\usepackage[labelfont=bf]{caption} % Bold figure x.x 
\captionsetup{width=0.9\textwidth}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\section{Results}
    1000 and 5000 SBC iterations are run for centered and reparameterised versions of the stocashtic volatility model. Rank statistics histograms are presented for the static parameters and the 1st, 500th and 1000th latent state parameters with 20 bins constructed for each parameter. The black horizontal line represents uniformity conditional on the number of bins and SBC iterations. The $\chi^2$ statistic is used to summarise the deviations from uniformity for all parameters since it is unrealistic to visually inspect all the histograms. The SBC algorithm is summarised below:

    1) for sim in 1000 iterations:
        2) Draw from prior: $\theta^{sim}\sim\pi (\theta)$
        3) Simulate dataset with 1000 observations: $y^{sim} \sim p(y|\theta^{sim})$
        4) Draw 999 posterior samples (post warmup) $\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})$
        5) Compute rank statistics $r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})$

    \subsection{Hamiltonian Monte Carlo}
    Figure \ref{fig:cphmc1k} and Figure \ref{fig:ncphmc1k} show the SBC results for the centered and reparameterisd model for HMC respectively. Both sets of results display relatively noisy rank statistic estimates with variation around the discrete uniform distribution.

    The centered parameterisation in particular displays a lack of uniformity for $\phi$ and $\sigma^2$. Rank statistics for $\sigma^2$ show two peaks of either end of the histogram and a left peak for $phi$. Posterior estimates of $\sigma^2$ are \textit{underdispersed} relative to the true parameter $\sigma^2_{sim}$ generated by the prior distribution. On the other hand, the posterior samples of $\phi$ are \textit{over estimating} the true parameter generated by the prior. This suggests the posterior geometry for these parameters (at least) are difficult to sample for the HMC algorithm.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_cp_1k.png}
        \caption{1000 SBC iterations for centered model using Hamiltoninan Monte Carlo. Histograms show noisy estimates around uniform distribution. In particular, $\sigma^2$ displays peaks on both ends of the histogram and $\phi$ shows a major on the left hand side. This suggests that the posterior samples of $\sigma^2$ are underdispersed relative to the prior distribution and posterior samples of $\phi$ are over estimating the true parameter.}
        \label{fig:cphmc1k}
    \end{figure}

    Estimates of $\sigma^2$ and $\phi$ are more uniform under the reparameterised model. Instead, we observe a peak on the left side for the 500th state parameter $h_{500}$. Overall the estimates for non centered parameterisation are more favourable.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_1k.png}
        \caption{1000 SBC iterations for non centered model using Hamiltonian Monte Carlo. The rank statistic distributions for $\sigma^2$ and $\phi$ have improved. There is some evidence of over estimation of the 500th state parameter.}
        \label{fig:ncphmc1k}
    \end{figure}

    The noisy histogram estimates can be improved by increasing the number of SBC iterations. Figure \ref{fig:cphmc5k} and Figure \ref{fig:ncphmc5k} show the results of the same parameters with 5000 SBC iterations. There is an improvement to the reparameterised model with all histograms looking approximately uniform with much smaller variation around the true uniform value.

    Centered parameteristion posess similar improvement for $\mu$ and the state parameters. However, there are still major peaks at both ends of $\sigma^2$ and $\phi$ suggesting underdispersion of the posterior estimates relative to the prior distribution.

    These results suggests that HMC overall struggles to sample from the centered parameteristion of the stochastic volatility model. The reparameterised model results in much better and consistent inference for this sampling strategy, conditional on the model.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_cp_5k.png}
        \caption{5000 SBC iterations for centered model using Hamiltoninan Monte Carlo. $\mu$ and the state parameters look approximately uniform. $\sigma^2$ still has major peaks on both the left and right hand side indicating underdispersion relative to the prior distribution. $\phi$ exhibits a similar shape to $\sigma^2$ although the peaks are much smaller.}
        \label{fig:cphmc5k}
    \end{figure}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_5k.png}
        \caption{5000 SBC iterations for reparameterised model using Hamiltoninan Monte Carlo. Parameters display uniformity with less noise around the true uniform value. This suggests that the HMC algorithm is returning the correct posterior estimates for the selected parameters.}
        \label{fig:ncphmc5k}
    \end{figure}

    Further to calibration, the HMC algorithm struggles to efficiently generate posterior samples from the centered parameterisation of stochastic volatility. Figure \ref{fig:hmcess} visualises the distribution of ESS for the static parameters after 5000 SBC iterations. The non centered model appears to generate independent samples much more efficiently than the centered model. This is consistent with HMC struggliing to sample overall from the posteriors of $\sigma^2$ and $\phi$.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ess.png}
        \caption{Effective sample sizes for static parameters after 5000 iterations of SBC and 1000 post warmup samples for Hamiltoninan Monte Carlo. The centered parameterisation struggles to generate independent samples for $\sigma^2$ and $\phi$. Non parameterisation demonstrates much more efficient sampling behaviour.}
        \label{fig:hmcess}
    \end{figure}

    \subsection{Gaussian Mixture Approximation}
    The Gaussian Mixture Approximation is run for 1000 and 5000 SBC iterations with the first 10,000 samples discarded as burnin (consistent with \citet{kim1998stochastic}). However, unlike KSC who take 750,000 posterior draws from the model, the number of posterior samples taken as part of the SBC procedure is 9,999. Taking 750,000 draws from the Gaussian mixture approximation would lead to issues given the disk space and compute time required from run SBC. Furthermore, the number of posterior samples is larger than the 999 taken for the SBC experiment run for Hamiltonian Monte Carlo. Earlier experiments attempted 999 post burn in draws for the Gaussian approximation, however there are concerns that such a short chain would not have converged onto the target posterior distribution. Therefore 9,999 draws was chosen as a compromise under both constraints.

    The centered Gaussian mixture approximation contains issues with calibration. Figure \ref{fig:cpksc1k} presents the results from SBC where there is a lack of uniformity across all parameters. In particular, there is a large peak on the left hand side for $\mu$ and the right hand hand side of $\phi$. This suggests over estimation and under estimation of the true parameter respectively. All other parameters do not have any key features worth noting beyond noisy estimates around the true discrete uniform value.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_cp_1k.png}
        \caption{1000 SBC iterations for centered Gaussian mixture approximation model. The rank statistics for all the selected parameters display a lack of uniformity. $\mu$ has a large left peak suggesting the algorithm is over estimating the true parameter. The converse can be said about $\phi$ which possesses a peak on the right hand side which implies underestimation of the true parameter.}
        \label{fig:cpksc1k}
    \end{figure}

    Increasing the SBC iterations to 5000 does not improve the results. $\sigma^2$ and the state parameters contain arguably less noisy estimates around the uniform distribution. However, the skewed shapes of $\mu$ and $\phi$ adds further evidence that the sampling strategy is not returning the correct posterior estimates. This can be seen in Figure \ref{fig:cpksc5k}.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_cp_5k.png}
        \caption{5000 SBC iterations for centered Gaussian mixture approximation model. Estimates for $\sigma^2$ and the state parameters have marginally improved around the uniformity. However, there is no improvement in $\mu$ or $\phi$, suggesting a lack of calibration from the MCMC algorithm.}
        \label{fig:cpksc5k}
    \end{figure}

    Furthermore, the ESS estimates for this model indicate a high degree of autocorrelation and difficulty in generating independent samples from the posterior distribution. The median ESS for $\mu$, $\phi$ and $\sigma^2$ are 559, 103, 39.5 respectively for 10,000 post burn in draws. Despite having 10 times the number of draws, the number of effectively independent samples is much smaller relative to HMC, suggesting that this bespoke sampling strategy is highly inefficient.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_ess.png}
        \caption{Effective sample sizes for static parameters after 5000 iterations of SBC and 1000 post burnin samples for the Gaussian Mixture model. The bespoke sampling strategy struggles to generate independent samples for all static parameters, suggesting a high degree of autocorrelation in the Markkov Chain.}
        \label{fig:kscess}
    \end{figure}

    The reparameterised model using the bespoke sampling strategy fails to generate sufficient results due to numerical errors and highly nonstationary chains. Figure X shows the traceplot from one simulation from the reparamterised model. The chain for $\phi$ is stuck to a value very close to 1 (unit root) and $\mu$ and $\sigma^2$ continuously increases in value until a numerical error is reached. Note that this occurs during the burnin phase of the markov chain. As a result, SBC is unable to be performed on this parameterisation of the model.

    There are a variety of reasons why this may happen. Perhaps the underlying software used to perform the sampling and kalman filter is not suitable for this paramterisation, or perhaps more broadly the MCMC strategy does not perform well with this parameterisation of the model. More work is required to diagnose the cause of this problem. Other avenues to improve the sampling of this model could be to rewrite the sampler from scratch or consider a different sampling method. Note that this is only one kind of reparamterisation and further research could explore other reparamterisations, such as non centered in scale.

    \subsection{Reweigthing rank statistics of Gaussian Mixture}
    \citet{kim1998stochastic} correct for approximation error in their method by using importance sampling weights. The reweighting procedure ensures that samples are drawn from the correct posterior density. This reweighting step is applied to the calculation of the expected value of the posterior density to get a weighted mean estimate from the samples drawn from the Gaussian mixture approximation. The rank statistic can be written as a function of the expected value, so the reweighting step can be applied to the calculation of the rank statistics to see if this correction improves the calibration of the sampling approach.

    Let $v(\theta, h)$ be the log weights defined as the log difference between the posterior densities of the true model with log chi squared errors $log\: g(\theta, h | y^{\ast}_t)$ and the Gaussian mixture model $log\:  k(\theta, h_t | y^{\ast}_t)$.

    $$
    \begin{aligned}
        v(\theta, h) = log\: g(\theta, h | y^{\ast}_t) - log\:  k(\theta, h_t | y^{\ast}_t) = const + log\: g(y|h) - log\: k(y^{\ast} | h)
    \end{aligned}
    $$

    Take the exponential and normalise the weights for the $l^{th}$ posterior draw (note the constants cancel out):

    $$
    \begin{aligned}
    w^l = \frac{exp(v(\theta, h)_l)}{\sum_i exp(v(\theta, h)_i)}
    \end{aligned}
    $$

    This gives the importance sampling weight. The expectation for any function of the posterior samples can be written as a function of the weights.

    $$
    \begin{aligned}
    S(\theta) &= 1[\theta_l < \theta^{sim}] \\
    E[S(\theta) | y] &= \int s(\theta) g(\theta | y) d\theta\\ 
    &= \frac{\int s(\theta)\times exp(v(\theta, h)) * k(\theta, h_t | y^{\ast}_t)d\theta d h}{\int exp(v(\theta, h)) * k(\theta, h_t | y^{\ast}_t)d\theta d h} 
    \end{aligned}
    $$

    Therefore, the expectation of the reweighted posterior samples is given by:

    $$
    \begin{aligned}
    E[S(\hat{\theta}) | y^{\ast}] = \sum_l^L S(\theta_l)w_l
    \end{aligned}
    $$

    The rank statistic can be rewritten as a function of the expectation and weights. This gives us the reweighted rank statistics.

    $$
    \begin{aligned}
    r = \sum_{l=1}^{L}1[\theta_{l} < \theta^{sim}] \approx  L\times E[S(\hat{\theta})] = L\times \sum_l^L S(\theta_l)w_l
    \end{aligned}
    $$

    The results from applying this reweighting step to the rank statistics are given on Figure \ref{fig:reweight1k}. There are no major improvements to the uniformity of rank statistics. Increasing the SBC iterations to 5000 as seen in Figure \ref{fig:reweight5k} also results in no major improvements. The shape of both sets of histograms is consistent with the shape of the unweighted rank statistics.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/sir_cp_1k.png}
        \caption{1000 SBC iterations for reweighted rank statistics from the Gaussian mixture approximation model. The shape of the histogram is consistent with the unweighted rank statistics.}
        \label{fig:reweight1k}
    \end{figure}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/sir_cp_5k.png}
        \caption{5000 SBC iterations for reweighted rank statistics from the Gaussian mixture approximation model. No major improvements are observed to the uniformity of rank statistics.}
        \label{fig:reweight5k}
    \end{figure}

    \subsection{Evaluating all models and parameters}
    As discussed, it is not practical to visually inspect the histograms of all parameters in the stochastic volatility model. Instead, the Chi squared statistic can be used to summarise the shape of the distribution and if desired, be used for hypothesis testing. However, isntead of performing hypothesis tests, another visual approach is taken to compare the calibration of the algorithms and parameters. 
    % However, performing all these tests suffers from the same overcommunication as presenting all the histograms at once.

    The distribution of chi squared statistics across all simulations and all parameters are plotted as histograms on Figure \ref{fig:allchisq}. A chi squared statistic of 0 implies the sample is exactly uniformly distributed. Any value away from 0 captures any variation or deviation away from uniformity. This provides a high level view of all parammeters - how close they are to uniformity and how close or far away they are relative to the other SBC results for other algorithms and parameterisations.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/dist_chisq.png}
        \caption{Distribution of chi squared statistics for all models (5000 SBC iterations).}
        \label{fig:allchisq}
    \end{figure}

    The distribution for HMC are clustered very closely to zero. Both results for the bespoke sampling method are distributed far from zero which suggests strong evidence of a lack of uniformity across all parameters. This is consistent with the inspected distriubtion of rank statistics in the previous sections.

@@

- Fair comparison between 999 samples and 9,999 samples.
